Spatio-angular fluorescence microscopy I. Basic theory

### 2018-11-03 papers/basic-theory/paper/paper.tex
Updates to section 2.

% Talon's custom libraries
\usepackage{empheq}
% \usepackage{bm}

% Talon's custom commands
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rb}{\mathbf{r}_b}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
\DeclareFontFamily{OT1}{pzc}{}
\DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling a microscope imaging an ensemble of in-focus
  fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular transfer function (SATF). We demonstrate our formalism by
  analyzing a single-view paraxial epi-fluorescence microscope without using the
  monopole or scalar approximations. We show that this microscope has an angular
  band limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with simple phantoms. Notably, we
  show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is recorded by paraxial epi-fluorescence microscopes. We
  discuss the implications of our analysis for all quantitative fluorescence
  microscopy studies and lay out a path towards a complete theory.
\end{abstract*}

@@

We begin in Section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently simplify our
analysis of spatio-angular microscopes. In Section \ref{sec:results} we use the
mathematical tools we developed in Section \ref{sec:theory} to analyze and
simulate a single-view paraxial epi-fluorescence microscope, and we demonstrate
our model with four simple phantoms. Finally, in Section \ref{sec:discussion} we
discuss the results and their broader implications.

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to draw new conclusions about all fluorescence microscopes. In
Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will show that the point spread
function and the optical transfer function are mappings between object and data
space expressed in two different bases. In Section \ref{sec:dipole} we will
extend the monopole imaging model to dipoles and examine the mapping in four
different bases.

@@

Next, we name the representations of our object and data in a specific basis. In
a delta function basis the object can be represented by a function $f(\ro)$
called the \textit{spatial monopole density}---the number of monopoles per unit
area at the two-dimensional position $\ro$. Similarly, in a delta function basis
the data can be represented by a function $g'(\rd')$ called the
\textit{irradiance}---the power received by a surface per unit area at position
$\rd'$. It may seem pedantic to emphasize that these representations are in a
delta function basis, but we highlight this point to make it clear that the
delta function basis is not special---we can choose a different basis at will.

@@

We have chosen to normalize the point spread function so that
\begin{align}
  \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}) = 1. \label{eq:norm}
\end{align}
The point spread function corresponds to a measurable irradiance, so it is
always real-valued and positive.

@@

The mapping between the object and the data in linear shift-invariant
microscopes takes a particularly simple form in a spatial harmonic basis. If we
apply the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bs{\nu}) = H(\bs{\nu})F(\bs{\nu}),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
\end{align}
the \textit{optical transfer function} as
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}).
\end{align}
The point spread function is normalized and real-valued, so we know that the
optical transfer function is normalized $H(0) = 1$ and conjugate-symmetric
$H(-\bs{\nu}) = H^*(\bs{\nu})$.

@@

\subsubsection{Monopole coherent transfer functions}

@@

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{spatio-angular dipole density}---the number of
dipoles at position $\ro{}$ per unit area oriented along $\so{}$ per unit solid
angle. Similar to the monopole case, we model the mapping between the object and
the irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so). 
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. Once again, many fluorescence
microscopes can be modeled as magnifiers with shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro, \so)f(\ro, \so). 
\end{align}

We define the same demagnified detector coordinate $\rd = \rd'/m$ and a new
normalization factor that corresponds to the total power incident on the
detector due to a spatial point source with an angularly uniform distribution of
dipoles
$P_\text{dip} = \int_{\mbb{S}^2}d\mb{s}\int_{\mbb{R}^2}d\mb{r}\, h'(m\mb{r},
\mb{s})$. We use these scaling factors to define the
\textit{orientation-dependent point spread function} as
\begin{align}
  h(\rd - \ro, \mb{s}) = \frac{h'(m[\rd - \ro], \mb{s})}{P_\text{dip}}, 
\end{align}
and the \textit{scaled irradiance} as 
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_\text{dip}}. 
\end{align}
With these definitions we can express the mapping between the object and the
data as an integral over the sphere with a spatial convolution at each point
\begin{align}
g(\rd{}) = \int_{\mbb{S}^2}d\so{}\int_{\mbb{R}^2}d\ro{}\, h(\rd{} -\ro{}, \so{})f(\ro, \so). \label{eq:odpsf}
\end{align}
Similar to the monopole case, we have chosen to normalize the orientation-dependent point spread function so that
\begin{align}
  \int_{\mbb{S}^2}d\mh{s}\int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \mh{s}) = 1. 
\end{align}
The orientation-dependent point spread function is a measurable quantity, so it
is real-valued and positive.

\subsubsection{Orientation-dependent optical transfer function}
We can make our first change of basis by applying the Fourier-convolution
theorem to Eq. \ref{eq:odpsf} which yields
\begin{align}
G(\bs{\nu}) = \int_{\mbb{S}^2}d\mh{s}\, H(\bs{\nu}, \mh{s})F(\bs{\nu}, \mh{s}) \label{eq:odotf},
\end{align}
where we define the \textit{orientation-dependent optical transfer function} as
  \begin{align}
  H(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
  \end{align}
and the \textit{spatial dipole spectrum} as
  \begin{align}
  F(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}). 
  \end{align}
  Since the orientation-dependent point spread function is normalized and real,
  we know that the orientation-dependent optical transfer function is normalized
  $\int_{\mbb{S}^2}d\mh{s}\, H(0, \mh{s}) = 1$ and conjugate-symmetric
  $H(-\bs{\nu}, \mh{s}) = H^*(\bs{\nu}, \mh{s})$.

@@

\subsubsection{Angular transfer function}
The spherical harmonics are another set of convenient basis functions. We can
change basis from spherical delta functions to spherical harmonics by applying
the generalized Plancharel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}{}\, p(\mh{s})q(\mh{s}) = \lmsum P_\ell^m Q_\ell^m, \label{eq:plan}
\end{align}
where $p(\mh{s})$ and $q(\mh{s})$ are arbitrary functions on the sphere and
\begin{align}
  P_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, p(\mh{s})Y_\ell^{m*}(\mh{s}), 
\end{align}
is the spherical Fourier transform of $p(\mh{s})$ and $Y_{\ell}^m(\mh{s})$ are
the spherical harmonic functions defined in Appendix \ref{sec:sph}. Eq.
\ref{eq:plan} expresses the fact that scalar products are invariant under a
change of basis (see Eq. 3.78 of \cite{barrett2004}). The left-hand side of Eq.
\ref{eq:plan} is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a
delta function basis and the right hand side is the scalar product of
$\mbb{L}_2(\mbb{S}^2)$ functions in a spherical harmonic function basis.
Applying Eq. \ref{eq:plan} to Eq. \ref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum H_\ell^m(\rd - \ro)F_\ell^m(\ro) \label{eq:atf-form}
\end{align}
where we have defined the \textit{angular transfer function} as
\begin{align}
  H_\ell^m(\mb{r}) = \int_{\mbb{S}^2}d\mh{s}\, h(\mb{r}, \mh{s})Y_{\ell}^{m*}(\mh{s}),\label{eq:atf-prep} 
\end{align}
and the \textit{angular dipole spectrum} as
\begin{align}
  F_\ell^m(\mb{r}) = \int_{\mbb{S}^2}d\mh{s}\, f(\mb{r}, \mh{s})Y_{\ell}^{m*}(\mh{s}).
\end{align}
Since the orientation-dependent point spread function is normalized and
real-valued, we know that the angular transfer function is normalized
$\int_{\mbb{R}^2}d\mb{r}\, H_0^0(\mb{r}) = 1$ and conjugate symmetric
$H_\ell^{-m}(\mb{r}) = (-1)^mH_\ell^{m*}(\mb{r})$.

@@

\subsubsection{Spatio-angular transfer function}
We can arrive at our final basis in two ways: by applying the generalized
Plancharel theorem for spherical functions to Eq. \ref{eq:odotf} or by applying
the Fourier convolution theorem to Eq. \ref{eq:atf-form}. We follow the
second path and find that
\begin{align}
G(\bs{\nu}) = \lmsum \msf{H}_\ell^m(\bs{\nu})\msf{F}_\ell^m(\bs{\nu}) \label{eq:saft},
\end{align}
where we have defined the \textit{spatio-angular transfer function} as
  \begin{align}
  \msf{H}_\ell^m(\bs{\nu}) &= \int_{\mbb{S}^2}d\mh{s}\, H(\bs{\nu}, \mh{s})Y_\ell^{m*}(\mh{s}),
  \end{align}
  and the \textit{spatio-angular dipole spectrum} as
  \begin{align}
  \msf{F}_\ell^m(\bs{\nu}) &= \int_{\mbb{S}^2}d\mh{s}\, F(\bs{\nu}, \mh{s})Y_\ell^{m*}(\mh{s}).
  \end{align}
  Since the orientation-dependent point spread function is normalized and
  real-valued, we know that the spatio-angular transfer function is normalized
  $\msf{H}_0^0(0) = 1$ and conjugate symmetric
  $\msf{H}_\ell^{-m}(-\bs{\nu}) = (-1)^m\msf{H}_\ell^{m*}(\bs{\nu})$.

  This basis is well suited for simulating and analyzing arbitrary samples
  because it exploits the sparsity of the imaging system. We will see this
  advantage explicitly in Section XXX when we calculate a specific
  spatio-angular transfer function and see that it has an angular band limit.

@@

\subsubsection{Dipole coherent transfer functions}
Two-dimensional Fourier transforms and spherical Fourier transforms are the only
operations required to change between the transfer functions. However, there is
a well-known alternative for calculating optical transfer functions in terms of
coherent transfer functions. The key idea is that the orientation-dependent
point spread function can always be written as the absolute square of a
vector-valued function called the \textit{orientation-dependent coherent spread
  function}
\begin{align}
  h(\mb{r}, \mh{s}) = |\mb{c}(\mb{r}, \mh{s})|^2. \label{eq:absquare}
\end{align}
Physically, this decomposition corresponds to the fact that the irradiance is
always the absolute square of a vector-valued electric field.

Since the orientation-dependent transfer function is given by the
two-dimensional Fourier transform of the orientation-dependent point spread
function
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\,h(\mb{r}, \mh{s})\,\text{exp}[-2\pi i\mb{r}\cdot\bs{\nu}],
\end{align}
we can plug in Eq. \ref{eq:absquare} and use the autocorrelation theorem to
rewrite as
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\bs{\tau}\,\mb{C}(\bs{\tau}, \mh{s})\mb{C}^\dagger(\bs{\tau} - \bs{\nu}, \mh{s}), 
\end{align}
where the \textit{orientation-dependent coherent transfer function}
$\mb{C}(\bs{\tau}, \mh{s})$ is the two-dimensional Fourier transform of the
orientation-dependent coherent spread function
\begin{align}
  \mb{C}(\bs{\tau}, \mh{s}) = \int_{\mbb{R}^2}d\mb{r}\, \mb{c}(\mb{r}, \mh{s})\,\text{exp}[-2\pi i\mb{r}\cdot\bs{\tau}].
\end{align}
The relationships between all six transfer function are shown in Figure
\ref{fig:transfer-functions}. Note that we could define coherent equivalents for
all four incoherent transfer functions, but we restrict ourselves to these two
since they are the useful for our calculations.

>>>>

% Talon's custom libraries
\usepackage{empheq}
\usepackage{bm}

% Talon's custom commands
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rb}{\mathbf{r}_b}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
% \DeclareFontFamily{OT1}{pzc}{}
% \DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling a microscope imaging an ensemble of in-focus
  fluorescent dipoles as a linear Hilbert-space operator with domain
  $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$, and we
  express the operator in terms of four different sets of object-space basis
  functions. We show that the operator takes a particularly convenient form when
  expressed in a basis of spatial and spherical harmonics---a form we call the
  spatio-angular dipole transfer function. We demonstrate our formalism by
  analyzing a single-view paraxial epi-fluorescence microscope without using the
  monopole or scalar approximations. We show that this microscope has an angular
  band limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with simple phantoms. Notably, we
  show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is recorded by paraxial epi-fluorescence microscopes. We
  discuss the implications of our analysis for all quantitative fluorescence
  microscopy studies and lay out a path towards a complete theory.
\end{abstract*}

@@

We begin in Section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently analyze
spatio-angular microscopes. In Section \ref{sec:results} we use the mathematical
tools we developed in Section \ref{sec:theory} to analyze and simulate a
single-view paraxial epi-fluorescence microscope, and we demonstrate our model
with four simple phantoms. Finally, in Section \ref{sec:discussion} we discuss
the results and their broader implications.

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to draw new conclusions about all fluorescence microscopes. In
Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will demonstrate the mapping in between
object and data in two different bases. In Section \ref{sec:dipole} we will
extend the monopole imaging model to dipoles and examine the mapping in four
different bases.

@@

Next, we name the representations of our object and data in a specific basis. In
a delta function basis the object can be represented by a function $f(\ro)$
called the \textit{monopole density}---the number of monopoles per unit area at
the two-dimensional position $\ro$. Similarly, in a delta function basis the
data can be represented by a function $g'(\rd')$ called the
\textit{irradiance}---the power received by a surface per unit area at position
$\rd'$. It may seem pedantic to emphasize that these representations are in a
delta function basis, but we highlight this point to make it clear that the
delta function basis is not special---we can choose a different basis at will.

@@

We have chosen to normalize the point spread function so that
\begin{align}
  \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}) = 1. \label{eq:norm}
\end{align}
The monopole point spread function corresponds to a measurable irradiance, so it
is always real-valued and positive.

@@

The mapping between the object and the data in linear shift-invariant
microscopes takes a particularly simple form in a spatial harmonic basis. If we
apply the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bs{\nu}) = H(\bs{\nu})F(\bs{\nu}),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
\end{align}
the \textit{monopole transfer function} as
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}).
\end{align}
The monopole point spread function is normalized and real-valued, so we know
that the monopole transfer function is normalized $H(0) = 1$ and
conjugate-symmetric $H(-\bs{\nu}) = H^*(\bs{\nu})$.

@@

We have used the term \textit{monopole transfer function} instead of the more
common \textit{optical transfer function}. We define the optical transfer
function as the mapping between an input irradiance spectrum and an output
irradiance spectrum in an optical system. With this definition the optical
transfer function is not necessarily the same as the monopole transfer function.
For example, consider a light sheet illumination scheme where a sheet of the
sample is illuminated and an orthogonal detection arm collects the emitted
fluorescence. Both the excitation path and the detection path can be described
by optical transfer functions, and the monopole transfer function is given by
the convolution of the two optical transfer functions. We will see that this is
a valuable distinction when we consider dipole transfer functions.

\subsubsection{Monopole coherent transfer functions}
Although the Fourier transform can be used to calculate the monopole transfer
function directly from the monopole point spread function, there is a well-known
alternative that exploits coherent transfer functions. The key idea is that the
monopole point spread function can always be written as the absolute square of a
scalar-valued \textit{monopole coherent spread function} defined as
\begin{align}
  h(\mb{r}) = |c(\mb{r})|^2. \label{eq:absquarescalar}
\end{align}
Physically, the monopole coherent spread function is the scalar-valued field on
the detector with appropriate scaling.

Since the monopole transfer function is given by the two-dimensional Fourier
transform of the monopole point spread function
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\,h(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bs{\nu}],
\end{align}
we can plug in Eq. \ref{eq:absquarescalar} and use the autocorrelation theorem to
rewrite the monopole transfer function as
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\bs{\tau}\, C(\bs{\tau})C^*(\bs{\tau} - \bs{\nu}), 
\end{align}
where $z^*$ denotes the complex conjugate of $z$, and we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the orientation-dependent coherent spread function
\begin{align}
  C(\bs{\tau}) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bs{\tau}].
\end{align}
Physically, the monopole coherent transfer function is the scalar-valued field
in a Fourier plane of the detector (often called the pupil or aperture plane)
with appropriate scaling. This provides a valuable shortcut for analyzing
arbitrary microscopes since it is usually straightforward to calculate the field
in the pupil. A typical approach for calculating the transfer functions is to
(1) calculate the field in the pupil, (2) scale the field to find the coherent
monopole transfer function, then (3) use the relationships in Figure
\ref{fig:monopole-transfer-functions} to calculate the other transfer functions.

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-transfer-functions/monopole-transfer-functions.pdf}
  \caption{The incoherent monopole transfer functions are related by a
    two-dimensional Fourier transform (right column). The coherent monopole
    transfer functions (left column) can be used to simplify the calculation of
    the remaining transfer functions.}
   \label{fig:monopole-transfer-functions}
 \end{figure}

@@

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{dipole density}---the number of dipoles at
position $\ro{}$ per unit area oriented along $\so{}$ per unit solid angle.
Similar to the monopole case, we model the mapping between the object and the
irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so). 
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. Once again, many fluorescence
microscopes can be modeled as magnifiers with shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro, \so)f(\ro, \so). 
\end{align}

We define the same demagnified detector coordinate $\rd = \rd'/m$ and a new
normalization factor that corresponds to the total power incident on the
detector due to a spatial point source with an angularly uniform distribution of
dipoles
$P_\text{dip} = \int_{\mbb{S}^2}d\mh{s}\int_{\mbb{R}^2}d\mb{r}\, h'(m\mb{r},
\mh{s})$. We use these scaling factors to define the
\textit{dipole point spread function} as
\begin{align}
  h(\rd - \ro, \so) = \frac{h'(m[\rd - \ro], \so)}{P_\text{dip}}, 
\end{align}
and the \textit{scaled irradiance} as 
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_\text{dip}}. 
\end{align}
With these definitions we can express the mapping between the object and the
data as
\begin{align}
g(\rd{}) = \int_{\mbb{S}^2}d\so{}\int_{\mbb{R}^2}d\ro{}\, h(\rd{} -\ro{}, \so{})f(\ro, \so). \label{eq:odpsf}
\end{align}

Similar to the monopole case, we have chosen to normalize the dipole point
spread function so that
\begin{align}
  \int_{\mbb{S}^2}d\mh{s}\int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \mh{s}) = 1. 
\end{align}
The dipole point spread function is a measurable quantity, so it is real-valued
and positive.

\subsubsection{Spatial dipole transfer function}
We can make our first change of basis by applying the Fourier-convolution
theorem to Eq. \ref{eq:odpsf} which yields
\begin{align}
G(\bs{\nu}) = \int_{\mbb{S}^2}d\mh{s}\, H(\bs{\nu}, \mh{s})F(\bs{\nu}, \mh{s}) \label{eq:odotf},
\end{align}
where we define the \textit{spatial dipole transfer function} as
  \begin{align}
  H(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}),
  \end{align}
  and the \textit{spatial dipole spectrum} as
  \begin{align}
  F(\bs{\nu}, \mh{s}) &= \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r}, \mh{s})\, \text{exp}(-2\pi i\mb{r}\cdot\bs{\nu}). 
  \end{align}
  Since the dipole point spread function is normalized and real, we know that
  the spatial dipole transfer function is normalized
  $\int_{\mbb{S}^2}d\mh{s}\, H(0, \mh{s}) = 1$ and conjugate-symmetric
  $H(-\bs{\nu}, \mh{s}) = H^*(\bs{\nu}, \mh{s})$.

@@

\subsubsection{Angular dipole transfer function}
The spherical harmonics are another set of convenient basis functions. We can
change basis from spherical delta functions to spherical harmonics by applying
the generalized Plancharel theorem for spherical functions
\begin{align}
  \ints{2}d\mh{s}{}\, p(\mh{s})q(\mh{s}) = \lmsum P_\ell^m Q_\ell^m, \label{eq:plan}
\end{align}
where $p(\mh{s})$ and $q(\mh{s})$ are arbitrary functions on the sphere, $P_\ell^m$ and $Q_\ell^m$ are their spherical Fourier transforms defined by
\begin{align}
  P_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, p(\mh{s})Y_\ell^{m*}(\mh{s}),
\end{align}
and $Y_{\ell}^m(\mh{s})$ are the spherical harmonic functions defined in
Appendix \ref{sec:sph}. Eq. \ref{eq:plan} expresses the fact that scalar
products are invariant under a change of basis (see Eq. 3.78 of
\cite{barrett2004}). The left-hand side of Eq. \ref{eq:plan} is the scalar
product of $\mbb{L}_2(\mbb{S}^2)$ functions in a delta function basis and the
right hand side is the scalar product of $\mbb{L}_2(\mbb{S}^2)$ functions in a
spherical harmonic function basis. Applying Eq. \ref{eq:plan} to Eq.
\ref{eq:odpsf} yields
\begin{align}
  g(\rd) = \lmsum H_\ell^m(\rd - \ro)F_\ell^m(\ro) \label{eq:atf-form}
\end{align}
where we have defined the \textit{angular dipole transfer function} as
\begin{align}
  H_\ell^m(\mb{r}) = \int_{\mbb{S}^2}d\mh{s}\, h(\mb{r}, \mh{s})Y_{\ell}^{m*}(\mh{s}),\label{eq:atf-prep} 
\end{align}
and the \textit{angular dipole spectrum} as
\begin{align}
  F_\ell^m(\mb{r}) = \int_{\mbb{S}^2}d\mh{s}\, f(\mb{r}, \mh{s})Y_{\ell}^{m*}(\mh{s}).
\end{align}
Since the dipole point spread function is normalized and real-valued, we know
that the angular dipole transfer function is normalized
$\int_{\mbb{R}^2}d\mb{r}\, H_0^0(\mb{r}) = 1$ and conjugate symmetric
$H_\ell^{-m}(\mb{r}) = (-1)^mH_\ell^{m*}(\mb{r})$.

@@

\subsubsection{Spatio-angular dipole transfer function}
We can arrive at our final basis in two ways: by applying the generalized
Plancharel theorem for spherical functions to Eq. \ref{eq:odotf} or by applying
the Fourier convolution theorem to Eq. \ref{eq:atf-form}. We follow the
second path and find that
\begin{align}
G(\bs{\nu}) = \lmsum \msf{H}_\ell^m(\bs{\nu})\msf{F}_\ell^m(\bs{\nu}) \label{eq:saft},
\end{align}
where we have defined the \textit{spatio-angular dipole transfer function} as
  \begin{align}
  \msf{H}_\ell^m(\bs{\nu}) &= \int_{\mbb{S}^2}d\mh{s}\, H(\bs{\nu}, \mh{s})Y_\ell^{m*}(\mh{s}),
  \end{align}
  and the \textit{spatio-angular dipole spectrum} as
  \begin{align}
  \msf{F}_\ell^m(\bs{\nu}) &= \int_{\mbb{S}^2}d\mh{s}\, F(\bs{\nu}, \mh{s})Y_\ell^{m*}(\mh{s}).
  \end{align}
  Since the dipole point spread function is normalized and real-valued, we know
  that the spatio-angular dipole transfer function is normalized
  $\msf{H}_0^0(0) = 1$ and conjugate symmetric
  $\msf{H}_\ell^{-m}(-\bs{\nu}) = (-1)^m\msf{H}_\ell^{m*}(\bs{\nu})$.

  This basis is well suited for simulating and analyzing arbitrary samples
  because it exploits the sparsity of the imaging system. We will see this
  advantage explicitly in Section XXX when we calculate a specific
  spatio-angular dipole transfer function and see that it has an angular band
  limit.

@@

\subsubsection{Dipole coherent transfer functions}
Similar to the monopole case, there is efficient way to calculate the transfer
functions using coherent transfer functions. The dipole point spread function
can always be written as the absolute square of a vector-valued function called
the \textit{dipole coherent spread function}
\begin{align}
  h(\mb{r}, \mh{s}) = |\mb{c}(\mb{r}, \mh{s})|^2. \label{eq:absquare2}
\end{align}
Physically, the dipole coherent spread function is an appropriately scaled
vector-valued electric field on the detector. We need a vector-valued coherent
transfer function since the polarization of the field plays a significant role
in dipole imaging.

Since the spatial dipole transfer function is given by the two-dimensional
Fourier transform of the dipole point spread function
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\mb{r}\,h(\mb{r}, \mh{s})\,\text{exp}[-2\pi i\mb{r}\cdot\bs{\nu}],
\end{align}
we can plug in Eq. \ref{eq:absquare2} and use the autocorrelation theorem to
rewrite the spatial dipole transfer function as 
\begin{align}
  H(\bs{\nu}) = \int_{\mbb{R}^2}d\bs{\tau}\,\mb{C}(\bs{\tau}, \mh{s})\mb{C}^\dagger(\bs{\tau} - \bs{\nu}, \mh{s}), 
\end{align}
where we have introduced the \textit{dipole coherent transfer function}
$\mb{C}(\bs{\tau}, \mh{s})$ as the two-dimensional Fourier transform of the
dipole coherent spread function
\begin{align}
  \mb{C}(\bs{\tau}, \mh{s}) = \int_{\mbb{R}^2}d\mb{r}\, \mb{c}(\mb{r}, \mh{s})\,\text{exp}[-2\pi i\mb{r}\cdot\bs{\tau}].
\end{align}
Physically, the dipole coherent transfer function corresponds to the
vector-valued electric field created by a dipole with oriented along $\mh{s}$ in
a Fourier plane of the detector with appropriate scaling. As with the monopole
case we can calculate the dipole-orientation-dependent fields in the pupil
plane, scale them appropriately to find the dipole coherent transfer function,
the use the relationships in Figure \ref{fig:transfer-functions} to calculate
the other transfer functions. Note that we could define coherent equivalents
for all four incoherent transfer functions, but we restrict ourselves to these
two since they are the most useful for our calculations.
